---
title: "Analysis of Transit Only Lane Violations for SFMTA"
author: Anthony Chau
date: 07-10-2018
output: github_document
---


```{r}
muni <- read.csv('muni_transitLanes.csv', strip.white = TRUE)
muni <- as.data.frame(muni)

# Check variables in muni
# str(muni)
# summary(muni)

```
# Cleaning the Data

First, we remove the Citation.Issue.Month column from the data frame since the
month is already included in the Citation.Issue.Date column. And, we remove the 
Last.Edited.Date column because this is not relevant for our analysis. Lastly,
we remove the Ticket.Number column.
```{r}
muni$Citaton.Issue.Month <- NULL
muni$Last.Edited.Date <- NULL
muni$Ticket.Number <-  NULL
```

Check for NA values
```{r}
# Check for NA values for each column
apply(muni, 2, function(x) any(is.na(x)))

```


Now, let's focus our attention on the date and time data within this dataset.

Initially, the citation issue date and time were stored as factor variables. We collapse date and time into a single column and convert it to a datetime object. This will make the data easier to work with in our analysis.
```{r}
# Convert citation dates and times to a Datetime object

muni$citation_date <- strptime(muni$Citation.Issue.Date, format ='%m/%d/%Y')


muni$citation_dateTime <- strptime(paste(muni$Citation.Issue.Date, 
                                            muni$Citation.Issue.Time), 
                                      format ='%m/%d/%Y %H:%M')

# Drop original citation issue date and time

muni$Citation.Issue.Date <- NULL
muni$Citation.Issue.Time <- NULL

```


We want to collaspe observations in the location column to only record a unique
factor level for a street name.
```{r}
library(stringr)

muni$Location <- as.character(muni$Location)

### Testing code

# Split location into street number and street name
# Logic: If a word boundary is proceeded by a digit, split the word

splitAddress <- strsplit(muni$Location, "(?<=\\d)\\b", perl=T)

# Create new matrix with street name column and street name column
y <- do.call(rbind, splitAddress)
y <- as.data.frame(y)


# Change column names
colnames(y) <- c('Street.Number', 'Street.Name')

# Make new column in muni df
muni$street_name <- y$Street.Name

# Check df
str(muni)

# Check unique street names
unique(muni$street_name)

```

# Clean up street names
```{r}
# To extend functionality, can make it so that any street type will be converted to 
# empty string

# Vector of street types for use with str_replace_all function
street_types <- c("ST$" = "", "BLVD$" = "", "TUNL$" = "", "Street" = "", "Ave" = "",
                  "Street$" = "")

# implement function with magarittr
library(magrittr)

locationName_clean <- function(location){
  location %>% 
  as.character() %>% 
  str_replace_all("[:punct:]", "") %>% 
  str_replace_all(street_types) %>% 
  str_to_title() %>% 
  trimws(which="both")
}

muni$street_name <- locationName_clean(muni$street_name)

# Check
unique(muni$street_name)


```


Clean up minor spelling mistakes
```{r}

# Clean up spelling and duplicates of street names
# Find more efficient way of doing this
# Only keep indicator of the street for the 'numerical streets' because San Francisco
# has both 2nd St and 2nd Ave
# Remove for the rest to keep it more clean
# We will determine the exact geographical location with longitude and latitude
# coordinates
muni$street_name[muni$street_name %in% c("Ofarrell", "Ofallell")] <- "O'Farrell"
muni$street_name[muni$street_name %in% c("4Th", "04Th", "O4th")] <- "4th St"
muni$street_name[muni$street_name %in% c("3Rd", "03Rd")] <- "3rd St"
muni$street_name[muni$street_name %in% c("22Nd")] <- "22nd St"
muni$street_name[muni$street_name %in% c("Po")] <- "Post"


# Check
unique(muni$street_name)
```




Work on this
```{r}
library(ggplot2)
library(dplyr)


# Convert citation date to date object to work in dplyr
muni$citation_date <- as.Date(muni$citation_date, format ='%m/%d/%Y')
muni$citation_dateTime <- as.Date(muni$citation_dateTime, format ='%m/%d/%Y  %H:%M') 

# Rename violations
muni <- muni %>% 
  rename(violation_type = Violation)

# Group together factor level with low counts

# Visualize number of violations by street name with barplot
streetBarPlot <- ggplot(data=muni, aes(x=street_name)) + 
  geom_bar(aes(fill=violation_type), color='black', position = "dodge") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

streetBarPlot
```

We find that most transit lane violations are concentrated on a few streets: Geary, Market, Mission, Folsom, Stockton, and Sutter. Almost all of these pass through downtown
San Francisco, which is plausible since many bus services aggregate in downtown. 

Many questions can be asked from this visualization. Is the high number of citations of these streets because of high bus and therefore high transit lane density in downtown San Francisco? 

Are drivers more likely to violate regulations in the downtown area because of how crowded and hectic streets can get during peak hours, which makes driver more 
short-tempered and willing to make shortcuts through transit lanes?

Or, are high incidence of transit lane violations on a few streets because muni drivers
are more short-tempered and willing to report violations in crowded downton SF?

Also, we have to take into account when each transit lane was implemented.



# Visualize number of violations with barplot
```{r}
# Visualize number of violations by street name with barplot
violations <- ggplot(data=muni, aes(x=violation_type)) + 
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

violations
```
Most violation types are towaway zone and towaway zone #1. There are few on sidewalk and OVR 18*C violations. As there are few no violations. Need more information on how specific violation types are defined.


Violations - Count Plot
```{r}
violationsCountPlot <- ggplot(data=muni)+ 
  geom_count(mapping = aes(x = violation_type, y = street_name)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


violationsCountPlot
```

Violations by Street Name - Tile Plot (possibly do a heat map with d3heatmap or heatmaply)

```{r}
muni %>% 
  count(violation_type, street_name) %>% 
  ggplot(mapping = aes(x = violation_type, y = street_name)) +
    geom_tile(mapping = aes(fill = n)) +   
    theme(axis.text.x = element_text(angle = 90, hjust = 1))




```



Violations by Street Name - Jitter plot
```{r}
violations_jitter <- ggplot(data = muni) +
  geom_jitter(mapping = aes(x = violation_type, y = street_name)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

violations_jitter
  
  
```







Violations - Pie Chart
```{r}
# Visualize number of violations with pie chart
pie <- ggplot(muni, aes(x = "", fill = factor(violation_type))) + 
  geom_bar(width = 1) +
  theme(axis.line = element_blank(), 
        plot.title = element_text(hjust=0.5)) + 
  labs(fill="class", 
       x=NULL, 
       y=NULL, 
       title="Pie Chart of Violation Type", 
       caption="Source: muni")
  
pie + coord_polar(theta = "y", start=0)

```



Time Series visualization test
Account for dates with no citation
Make dates with sequence function between earliest date and latest date
# try missing_values <- date[!date in% sequence]
```{r}
library(dplyr)

# Create new column which stores how many citations were issued that day and build 
# time series object from that column

# Convert citation date to date object to work in dplyr
muni$citation_date <- as.Date(muni$citation_date, format ='%m/%d/%Y')
muni$citation_dateTime <- as.Date(muni$citation_dateTime, format ='%m/%d/%Y  %H:%M') 


# Group and count observations by date 
muni <- 
  muni %>%
  arrange(citation_date) %>%
  group_by(citation_date) %>%
  mutate(citation_count = n())

# Try n = n()
# Count missing values: (sum(is.na(x))) or set na.rm=TRUE

# Pull out citation_count vector
citation_count <- 
  muni %>%
  pull(citation_count)

# Time series
# More research on dealing with regular data (almost daily) and spanning across
# many years
citation_ts <- ts(citation_count, frequency = 7, start= c(2008,2))

plot.ts(citation_ts)



```

