---
title: "Analysis of Transit Only Lane Violations for SFMTA"
author: Anthony Chau
date: 07-10-2018
output: 
  github_document:
    toc: true
    toc_depth: 2
always_allow_html: yes
---

```{r, message=FALSE, results='hide', warning = FALSE}
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyverse)
library(leaflet)
library(plotly)
library(lubridate)
```


# Read in data
```{r read_data, results = 'hide', message=FALSE}
# Use read_csv from readr package (Faster, more reproducible, tibbles)
muni <- read_csv('muni_transitLanes.csv', trim_ws = TRUE, na = c("", "NA"))
muni <- as.data.frame(muni)

# Check variables in muni
str(muni)

```
# Cleaning the Data

First, we remove the Citation.Issue.Month column from the data frame since the
month is already included in the Citation.Issue.Date column. And, we remove the 
Last.Edited.Date and Ticket Number column because the columns are not relevant for our analysis.
```{r remove_columns}

muni <- muni %>% 
  select(-`Citaton Issue Month`, -`Last Edited Date`, -`Ticket Number`)

```

## Missing Values

We have many missing values scattered across the columns in the muni data frame. More specifcally, there are missing values in the 'Amount Due', 'Suspend Code', 'Suspend Process Date', 'Suspend Until Date', 'Disposition Code', and 'Geom' columns.

we have many missing values in the Disposition Code column but few missing values for the 'Amount Due' and Geom' column. Also, there are about an equal amount of missing values for the 'Suspend Code', 'Suspend Process Date, and 'Suspend Until Date'. This makes sense intuitively because the columns are related to each other. In fact the 'Suspend Process Date' and 'Suspend Until Date' have the same number of missing values. 
```{r missing_values}
# Check for NA values for each column

apply(muni, 2, function(x) any(is.na(x)))

# find total number of missing values for columns with missing values
compute_missing <-  function(missing_column){
  cat("Total missing values:", sum(is.na(missing_column)), "\n")
}

compute_missing(muni$`Amount Due`)
compute_missing(muni$`Suspend Code`)
compute_missing(muni$`Suspend Process Date`)
compute_missing(muni$`Suspend Until Date`)
compute_missing(muni$`Disposition Code`)
compute_missing(muni$Geom)

```

## Impute Missing Valuess
```{r}

```


## Date and Time

Now, let's focus our attention on the date and time data within this dataset.

Initially, the citation issue date and time were stored as factor variables. We collapse date and time into a single column and convert it to a datetime object. This will make the data easier to work with in our analysis.

We also want to analyze citation counts in a smaller time frame, so we separate the citation_date column into three columns for the month, date, and year.
```{r dateTime}
# Convert all values in the original citation column into a Date object
muni$citation_date <- as.Date(muni$`Citation Issue Date`, format ="%m/%d/%Y")


# Combine date and time columns into a citation_dateTime column
muni <-  muni %>% 
  unite(citation_dateTime, `Citation Issue Date`, `Citation Issue Time`, sep = " ")

# Convert citation_dateTime to a POSIXct object
muni$citation_dateTime <- mdy_hms(muni$citation_dateTime)

# Separate date column
muni <- muni %>% 
  separate(citation_date, into = c("year", "month", "day"), sep = "-", remove = FALSE)

# Convert month, day, and year columns into the integer type
muni$month <- as.integer(muni$month)

muni$day <- as.integer(muni$day)

muni$year <- as.integer(muni$year)

```


## Spatial Location 
```{r}
# Split Geom column into Longitude and Latitude column
muni <-  muni %>% 
  separate(Geom, into = c("latitude", "longitude"), sep = ",")

# Clean up longitude and latitude column (remove parenthesis)
muni$latitude <- muni$latitude %>% 
  str_replace("^\\(", "")

muni$longitude <-  muni$longitude %>% 
  str_replace("\\)$", "")

```


## Extract unique street names from location column 

In the original datset, the location column provided both the street number and street name. Hence, we have an extremely high amount of factor levels for the location column. So, we create a new column which consists of only the street name where the violation occured. This step condenses down the number of factor levels to the total number of unique street names.
```{r street_names}

# Good functional programming practice:
# Not advisable to change global variables in functions
# Instead, we write a function with a more explicit return value and reassign the result
# outside the function (global environment)

# Separate Location column
muni <- muni %>% 
    separate(Location, c("street_number", "street_name"), sep = "(?<=\\d)\\b")

# Check street names
unique(muni$street_name)


### Explanation of regular expression

# Split location into street number and street name
# Logic: If a word boundary is proceeded by a digit, split the word



```

## Clean up street names

In the newly created street_name column, we remove unnecessary puncation or white space within the different factor levels and remove the street type from each factor level. Lastly, we convert each street name to title style capitalization.
```{r}
# To extend practicality and to simplify our string data, any street type will be converted to empty string

# Vector of street types for use with str_replace_all function
street_types <- c("ST$" = "", "BLVD$" = "", "TUNL$" = "", "Street" = "", "AVE" = "",
                  "STREET$" = "")


clean_streetName <- function(streetName_column){
  streetName_column <- streetName_column %>% 
  as.character() %>% 
  str_replace_all(c("[:punct:]" = "", street_types)) %>% 
  str_to_title() %>% 
  trimws(which = "both")
}

muni$street_name <- clean_streetName(muni$street_name)


# Double check
unique(muni$street_name)


```


## Clean up minor spelling mistakes

Some of the street names in the dataset are misspelled. We manually correct these
mistakes by reassigning observations with misspelling to the correct spelling.

For this analysis, I have decided to remove the street type indicator from non-duplicate
street names. Beacuse of this, only numerical street names, such as 4th or 22nd, have
street type indicators.
```{r}

# Clean up spelling and duplicates of street names
# Find more efficient way of doing this
# Only keep indicator of the street for the 'numerical streets' because San Francisco
# has both 2nd St and 2nd Ave
# Remove for the rest to keep it more clean
# We will determine the exact geographical location with longitude and latitude
# coordinates
muni$street_name[muni$street_name %in% c("Ofarrell", "Ofallell")] <- "O'Farrell"
muni$street_name[muni$street_name %in% c("4Th", "04Th", "O4th")] <- "4th St"
muni$street_name[muni$street_name %in% c("3Rd", "03Rd")] <- "3rd St"
muni$street_name[muni$street_name %in% c("22Nd")] <- "22nd St"
muni$street_name[muni$street_name %in% c("Po")] <- "Post"

# Change street_name to factor variable
muni$street_name <- factor(muni$street_name)

# Check
unique(muni$street_name)
```

## Rename columns
```{r}

muni <- muni %>% 
  rename(ID = "Object ID", violation_type = "Violation", fine_amount = "Fine Amount")

names(muni)

```

## Rearrange columns

```{r}
muni <- muni %>% 
  select(ID, month, day, year, street_name, 
         violation_type, latitude, longitude, 
         fine_amount, citation_dateTime, everything())
```


# Data Visualization

## Transit lane citations count from 2008-2015

First, we condense down the muni data frame by selecting only the street_name and citation_dateTime column. We will pass in this condensed data frame to ggplot2 to create visualizations.

Then, we make a frequency plot with ggplot2 to display the distribution of citation counts over time. However, this plot is difficult to interpret when our binwidth is set equal to 1 day because of the great volume of observations every day. We will choose a bigger binwidth for more useful visualizaion
```{r, fig.keep = 'last'}
# select citation_dateTime and street_name columns only
citations_dateTime <- muni %>% 
  select(street_name, citation_dateTime)

# Citation count by day from 2008 to 2014; 86400 secs = 1 day

citations_dateTime %>% 
  ggplot(aes(citation_dateTime)) +
  geom_freqpoly(binwidth = 86400, color = "firebrick", alpha = 0.3) +
  labs(x = "", title = "Transit Lanes Citation Count in San Francisco") + 
  theme_minimal()

```

## Transit lane citations count from 2008-2015 with increased binwidth

We increase the bindwidth to be approximately 7 days and put breaks in the dates every 6 months. This allows us to see a clearer general trend without the unnecessary noise.

From the graph, transit lane citations were increasing in general, with peak levels around September 2012. There is a cylical pattern in the data, with peaks and troughs scattered throughout the curve.
```{r warning= FALSE, fig.keep = 'last'}
citations_timeSeries <- citations_dateTime %>% 
  ggplot(aes(citation_dateTime)) +
  # take average days in a month to be 30 and multiple by the number of seconds in a day
  geom_freqpoly(binwidth = 86400*7, aes(color = factor(year(citation_dateTime)))) +
  labs(x = "", y = "Citation Count", 
       title = "Transit Lane Citation Counts in San Francisco") +
  scale_x_datetime(date_breaks = "6 months", date_labels = "%b - %Y") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, 
                                   margin = margin(0, 0, 15, 0)),
        plot.title = element_text(size = 15, face = "bold", 
                                  margin = margin(0, 0, 10, 0)),
        legend.title = element_blank())

citations_timeSeries

# plotly plot
ggplotly(citations_timeSeries, tooltip = c("y"))



```


## Visualizing Transit lane citations by street name. 

We will focus on the streets with relatively high amounts of citations. Namely, we will show distributions for Geary, Mission, O'Farrell, Post, and Sutter.

```{r}
# Number of citations across streets
muni %>% 
  group_by(street_name) %>% 
  summarise(n())
```


## Transit Lane Citations by Street Name

We display citation counts for different streets in San Francisco from 2008 to 2015. 

Some key observations from the plot:

1. Citation counts increase to a peak and then decreases after the peak for many streets. The initial upward trend of citation counts can be attributed to stricter enforcement when the transit lanes were first implemented. Then, after the behaviors of drivers have adjusted, strict enforcement decreased.

2. Sutter and Mission St have a disproportionately high amount of citation counts compared to the other streets

3. Citation counts for Mission St have two peaks. The two peaks may have been because  new transit lanes were rolled out in different phases.

4. Post and Sutter St sustain higher citation counts after hitting the peak compared to other streets

```{r, fig.keep = 'last'}
citations_groupedByStreet <- muni %>% 
  group_by(street_name, year) %>% 
  mutate(count = n())

citations_byStreet <- citations_groupedByStreet %>% 
  filter(street_name %in% c("Geary", "Mission", "O'Farrell", "Post", "Sutter")) %>% 
  ggplot(aes(street_name)) +
  geom_bar(color = "black", mapping = aes(fill = street_name)) +
  labs(x ="", y = "Citation Count", 
       title = "Citations count across different streets") + 
  facet_grid(cols = vars(year), rows = vars(street_name), 
             scales = "free_y") +
  theme_dark() + 
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(margin = margin(10, 0, 0, 0)), 
        axis.title.y = element_text(margin = margin(0, 10, 0, 0)),
        plot.title = element_text(face = "bold"),
        panel.spacing.y = unit(.5, "cm"),
        panel.grid.major = element_blank(),
        panel.border = element_rect(linetype = "dashed", fill = NA),
        strip.background.x = element_rect(colour = "white", fill = "gray"),
        strip.background.y = element_rect(colour = "white", fill = "firebrick"),
        strip.text.x = element_text(colour = "white", face = "bold"),
        panel.background = element_rect(fill = "white", color = "gray50"))

citations_byStreet

# glitchy
ggplotly(citations_byStreet, tooltip = "count")


``` 



## Transit Lane Citations by hour

Transit lane citations occur most frequently during afternoon rush hour, with a peak count between 4-5PM. There is a big difference in citation counts before the peak hour and citation counts do not gradually increase as time gets closer to the peak hour. Instead, citation counts stays fairly flat. 
```{r, fig.keep = 'last'}
citations_byHour <- muni %>% 
  ggplot(aes(hour(citation_dateTime))) +
  geom_bar(stat = "count", fill = "#39a78e") +
  theme_minimal() +
  labs(x = "Hour", y = "Total Count", title = "Transit Lane Citations by Hour") + 
  theme(axis.title.y = element_text(margin = margin(0, 10, 0, 0))) + 
  scale_x_continuous(breaks = seq(0, max(hour(muni$citation_dateTime)), by = 1))

citations_byHour

```


## Transit Lane Citations by year

Citations counts have been increasing from 2008 and onwards, peaking in 2012, but maintaining relatively high counts afterwards. New transit lanes being rolled out and the subsequent inital strict enforcement is a big reason for this increasing trend. More data from 2015 onwards is necessary to visualize future trends in citation counts.
```{r, fig.keep = 'last'}
citations_byYear <- muni %>% 
  ggplot(aes(year(citation_dateTime))) +
  geom_bar(stat = "count", fill = "#39a78e") +
  theme_minimal() +
  labs(x = "Year", y = "Total Count", title = "Transit Lane Citations by Year") + 
  theme(axis.title.y = element_text(margin = margin(0, 10, 0, 0))) + 
  scale_x_continuous(breaks = seq(min(year(muni$citation_dateTime)),
                                  max(year(muni$citation_dateTime)), by = 1))

citations_byYear
```



## Transit Lane Citations by Month

Citation counts peak around the month of July and are in general highest during Fall months (August, Septemeber, October). This makes sense because these months are when students go back to school and when people in the workforce have come back from summer vacation, increasing people and traffic flow throughout the city. In contrast, citation counts are lowest around the holidays and New Years, when many people leave the city for the holidays and are off from work.
```{r}
citations_byMonth <- muni %>% 
  ggplot(aes(month(citation_dateTime))) +
  geom_bar(stat = "count", fill = "#39a78e") +
  theme_minimal() +
  labs(x = "Month", y = "Total Count", title = "Transit Lane Citations by Month") + 
  theme(axis.title.y = element_text(margin = margin(0, 10, 0, 0)),
        axis.title.x = element_text(margin = margin(10, 0, 0, 0))) + 
  scale_x_continuous(breaks = seq(1, max(month(muni$citation_dateTime)), by = 1))

citations_byMonth

```



## Transit Lane Citations by Weekday
```{r}

```


## Transit Lane Citations by Violation Type

We have low observations counts for the following categories of violations: 'NO VIOL', 'ON SIDEWLK', 'OVR 18" C', and 'PK FR LB'. So, we remove these categories and the unique observations to make out analysis cleaner.
```{r}

# Counts of violation type
table(muni$violation_type)

unique(muni$violation_type)

# Find the 8 observations with the 4 violation categories and delete them from data frame

muni <- muni[!muni$violation_type %in% c('NO VIOL', 'ON SIDEWLK', 'OVR 18 " C', 'PK FR LN'),]

# Check violation categories
table(muni$violation_type)

# Check total number of observations - we orignially had 17,178 observations
dim(muni)


```

Most violation types are towaway zone #1, DNL PARK, and BUS ZONE. There are few observations of the on sidewalk, OVR 18*C violations, and no violation type. We need more information on how specific violation types are defined.

```{r, fig.keep = 'last'}

# Visualize number of violations by street name with barplot
citations_byType <- ggplot(data=muni, aes(x=violation_type)) + 
  geom_bar(fill = "firebrick") +
  theme_minimal() +
  labs(title = "Citation counts by type of violation", x = "Violation Type") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.title.x = element_text(margin = margin(10, 0, 0, 0)),
        axis.title.y = element_text(margin = margin(0, 10, 0, 0)))

citations_byType

```


# Geospatial mapping 

## Visualize transit lane violations with Leaflet
```{r}

muni$latitude <- as.numeric(muni$latitude)

muni$longitude <- as.numeric(muni$longitude)

leaflet(data = muni) %>% 
  addProviderTiles("Stamen.TonerLite") %>% 
  addCircleMarkers(~longitude, ~latitude, popup = "Transit Violation", 
                   label = "Got ya!",
                   clusterOptions = markerClusterOptions())
```



# Extensions/Further Work

1. Integrate SFMTA on-time arrival/departure and average delays for bus routes using transit only lanes -> How effective are transit-only lanes? How do we measure this?

2. Data is inconsistent and not frequently maintained/updated -> difficult to fully explore and understand the data

3. Research more on SFMTA and even city-wide intiatives to better understand spikes and dips in transit lane citations

4. Are drivers more likely to violate regulations in the downtown area because of how crowded and hectic streets can get during peak hours, which makes driver more 
short-tempered and willing to make shortcuts through transit lanes?

5. Are high incidence of transit lane violations on a few streets because muni drivers
are more short-tempered and willing to report violations in crowded downton SF?





