---
title: "Analysis of Transit Only Lane Violations for SFMTA"
author: Anthony Chau
date: 07-10-2018
output: github_document
---


```{r}
muni <- read.csv('muni_transitLanes.csv', strip.white = TRUE)
muni <- as.data.frame(muni)

# Check variables in muni
# str(muni)
# summary(muni)

```
# Cleaning the Data

First, we remove the Citation.Issue.Month column from the data frame since the
month is already included in the Citation.Issue.Date column. And, we remove the 
Last.Edited.Date column because this is not relevant for our analysis. Lastly,
we remove the Ticket.Number column.
```{r}
muni$Citaton.Issue.Month <- NULL
muni$Last.Edited.Date <- NULL
muni$Ticket.Number <-  NULL
```

Check for NA values
```{r}
# Check for NA values for each column
apply(muni, 2, function(x) any(is.na(x)))

```


Now, let's focus our attention on the date and time data within this dataset.

Initially, the citation issue date and time were stored as factor variables. We collapse date and time into a single column and convert it to a datetime object. This will make the data easier to work with in our analysis.
```{r}
# Convert citation dates and times to a Datetime object

muni$citation_date <- strptime(muni$Citation.Issue.Date, format ='%m/%d/%Y')


muni$citation_dateTime <- strptime(paste(muni$Citation.Issue.Date, 
                                            muni$Citation.Issue.Time), 
                                      format ='%m/%d/%Y %H:%M')

# Drop original citation issue date and time

muni$Citation.Issue.Date <- NULL
muni$Citation.Issue.Time <- NULL

```


We want to collaspe observations in the location column to only record a unique
factor level for a street name.
```{r}
library(stringr)

muni$Location <- as.character(muni$Location)

### Testing code

# Split location into street number and street name
# Logic: If a word boundary is proceeded by a digit, split the word

splitAddress <- strsplit(muni$Location, "(?<=\\d)\\b", perl=T)

# Create new matrix with street name column and street name column
y <- do.call(rbind, splitAddress)
y <- as.data.frame(y)


# Change column names
colnames(y) <- c('Street.Number', 'Street.Name')

# Make new column in muni df
muni$street_name <- y$Street.Name

# Check df
str(muni)

# Check unique street names
unique(muni$street_name)

```

# Clean up street names
```{r}
# To extend functionality, can make it so that any street type will be converted to 
# empty string

# Vector of street types for use with str_replace_all function
street_types <- c("ST$" = "", "BLVD$" = "", "TUNL$" = "", "Street" = "", "Ave" = "")

# implement function with magarittr
library(magrittr)

locationName_clean <- function(location){
  location %>% 
  as.character() %>% 
  str_replace_all("[:punct:]", "") %>% 
  str_replace_all(street_types) %>% 
  str_to_title() %>% 
  trimws(which="both")
}

muni$street_name <- locationName_clean(muni$street_name)

# Check
unique(muni$street_name)


```


Clean up minor spelling mistakes
```{r}

# Clean up spelling and duplicates of street names
# Find more efficient way of doing this
# Only keep indicator of the street for the 'numerical streets' because San Francisco
# has both 2nd St and 2nd Ave
# Remove for the rest to keep it more clean
# We will determine the exact geographical location with longitude and latitude
# coordinates
muni$street_name[muni$street_name %in% c("Ofarrell", "ofallell")] <- "O'Farrell"
muni$street_name[muni$street_name %in% c("4Th", "04Th", "04th")] <- "4th St"
muni$street_name[muni$street_name %in% c("3Rd", "03Rd")] <- "3rd St"
muni$street_name[muni$street_name %in% c("22Nd")] <- "22nd St"
muni$street_name[muni$street_name %in% c("Po")] <- "Post"


# Check
unique(muni$street_name)
```




Work on this
```{r}
library(ggplot2)

# Group together factor level with low counts

# Visualize number of violations by street name with barplot
streetBarPlot <- ggplot(data=muni, aes(x=street_name)) + 
  geom_bar(aes(fill=Violation), color='black') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

streetBarPlot

# Visualize number of violations by street name with barplot
violations <- ggplot(data=muni, aes(x=Violation)) + 
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

violations

# Visualize number of violations with pie chart
pie <- ggplot(muni, aes(x = "", fill = factor(Violation))) + 
  geom_bar(width = 1) +
  theme(axis.line = element_blank(), 
        plot.title = element_text(hjust=0.5)) + 
  labs(fill="class", 
       x=NULL, 
       y=NULL, 
       title="Pie Chart of Violation Type", 
       caption="Source: muni")
  
pie + coord_polar(theta = "y", start=0)


violations

```

Time Series visualization test
Account for dates with no citation
Make dates with sequence function between earliest date and latest date
# try missing_values <- date[!date in% sequence]
```{r}
library(dplyr)

# Create new column which stores how many citations were issued that day and build 
# time series object from that column

# Convert citation date to date object to work in dplyr
muni$citation_date <- as.Date(muni$citation_date, format ='%m/%d/%Y')
muni$citation_dateTime <- as.Date(muni$citation_dateTime, format ='%m/%d/%Y  %H:%M') 


# Group and count observations by date 
muni <- 
  muni %>%
  arrange(citation_date) %>%
  group_by(citation_date) %>%
  mutate(citation_count = n())

# Try n = n()
# Count missing values: (sum(is.na(x))) or set na.rm=TRUE

# Pull out citation_count vector
citation_count <- 
  muni %>%
  pull(citation_count)

# Time series
# More research on dealing with regular data (almost daily) and spanning across
# many years
citation_ts <- ts(citation_count, frequency = 7, start= c(2008,2))

plot.ts(citation_ts)



```

