---
title: "Analysis of Transit Only Lane Violations for SFMTA"
author: Anthony Chau
date: 07-10-2018
output: html_notebook
---


```{r}
muni <- read.csv('muni_transitLanes.csv', strip.white = TRUE)
muni <- as.data.frame(muni)

# Check variables in muni
str(muni)
summary(muni)

```
# Cleaning the Data

First, we remove the Citation.Issue.Month column from the data frame since the
month is already included in the Citation.Issue.Date column. And, we remove the 
Last.Edited.Date column because this is not relevant for our analysis.
```{r}
muni$Citaton.Issue.Month <- NULL
muni$Last.Edited.Date <- NULL
```

Now, let's focus our attention on the date and time data within this dataset.

Initially, the citation issue date and time were stored as factor variables. We collapse date and time into a single column and convert it to a datetime object. This will make the data easier to work with in our analysis.
```{r}
# Convert citation dates and times to a Datetime object

muni$Citation.Date <- strptime(paste(muni$Citation.Issue.Date, 
                                            muni$Citation.Issue.Time), 
                                      format ='%m/%d/%Y %H:%M')

# Drop original citation issue date and time

muni$Citation.Issue.Date <- NULL
muni$Citation.Issue.Time <- NULL

```


We want to collaspe observations in the location column to only record a unique
factor level for a street name.
```{r}
library(stringr)

muni$Location <- as.character(muni$Location)

muni$Location <- str_replace_all(muni$Location, "[:punct:]", "")

### Testing code

# Split location into street number and street name
# Logic: If a word boundary is proceeded by a digit, split the word

splitAddress <- strsplit(muni$Location, "(?<=\\d)\\b ", perl=T)

# Create new matrix with street name column and street name column
y <- do.call(rbind, splitAddress)
y <- as.data.frame(y)


# Change column names
colnames(y) <- c('Street.Number', 'Street.Name')

# Make new column in muni df
muni$Street.Name <- y$Street.Name

# Check df
str(muni)

# Check unique street names
unique(muni$Street.Name)

# Trim leading white space
trim.leading <- function (x)  sub("^\\s+", "", x)
muni$Street.Name <- trim.leading(muni$Street.Name)

# Refactor duplicate/misspelled streets
# Find more efficient way of doing this
muni$Street.Name[muni$Street.Name %in% c("GEARY", "GEARY ST")] <- "GEARY BLVD"
muni$Street.Name[muni$Street.Name %in% c("OFARRELL", "OFALLELL", 
                                         "OFARRELL ST", 
                                         "OFARRELL STREET")] <- "O'FARRELL ST"
muni$Street.Name[muni$Street.Name %in% c("04TH ST", "04TH  ST",
                                         "O4TH STREET")] <- "4TH ST"
muni$Street.Name[muni$Street.Name %in% c("03RD ST")] <- "3RD ST"
muni$Street.Name[muni$Street.Name %in% c("MISSION")] <- "MISSION ST"
muni$Street.Name[muni$Street.Name %in% c("MARKET", "566MARKET ST")] <- "MARKET ST"
muni$Street.Name[muni$Street.Name %in% c("STOCKTON", 
                                         "STOCKTON TUNL")] <- "STOCKTON ST"
muni$Street.Name[muni$Street.Name %in% c("SACRAMENTO")] <- "SACRAMENTO ST"
muni$Street.Name[muni$Street.Name %in% c("POST")] <- "POST ST"
muni$Street.Name[muni$Street.Name %in% c("KEARNY")] <- "KEARNY ST"
muni$Street.Name[muni$Street.Name %in% c("SUTTER")] <- "SUTTER ST"
muni$Street.Name[muni$Street.Name %in% c("CLAY")] <- "CLAY ST"
muni$Street.Name[muni$Street.Name %in% c("MAIN")] <- "MAIN ST"
muni$Street.Name[muni$Street.Name %in% c("FOLSOM")] <- "FOLSOM ST"
muni$Street.Name[muni$Street.Name %in% c("TRUMBWELL")] <- "TRUMBWELL ST"
muni$Street.Name[muni$Street.Name %in% c("WILDE")] <- "WILDE ST"

unique(muni$Street.Name)


```

