---
title: "Analysis of Transit Only Lane Violations for SFMTA"
author: Anthony Chau
date: 07-10-2018
output: html_notebook
---


```{r}
muni <- read.csv('muni_transitLanes.csv')

# Convert muni object to a data frame class
muni <- as.data.frame(muni)

# Check variables in muni
str(muni)
summary(muni)

```
# Cleaning the Data

First, we remove the Citation.Issue.Month column from the data frame since the
month is already included in the Citation.Issue.Date column.
```{r}
muni$Citaton.Issue.Month <- NULL
```

Now, let's focus our attention on the date and time data within this dataset.

Initially, the citation issue date and time were stored as factor variables. We collapse date and time into a single column and convert it to a datetime object. This will make the data easier to work with in our analysis.
```{r}
# Convert citation dates and times to a Datetime object

muni$Citation.Date <- strptime(paste(muni$Citation.Issue.Date, 
                                            muni$Citation.Issue.Time), 
                                      format ='%m/%d/%Y %H:%M')

# Drop original citation issue date and time

muni$Citation.Issue.Date <- NULL
muni$Citation.Issue.Time <- NULL

```


